Silent pessimists: researchers' perceptions of malaria eradication and implications for health spending

### 2017-12-27 paper.Rmd
update

---
title: Silent pessimists - researchers' perceptions of malaria eradication and implications for health spending
bibliography: library.bib
output: 
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: preamble.sty
---

@@

Quantifying an event's probability and time frame is essential for calculating its expected value. In the case of global malaria eradication, uncertainty regarding feasibility makes it difficult for policymakers and public health practitioners to make fully informed decisions. The opportunity cost of investments in eradication-specific interventions can be high, particularly in contexts with other urgent health priorities. In a systematic survey of malaria researchers, we query perceptions regarding the likelihood and time frame of eradication, as well as the perceived chief areas needed for improvement in order for eradication to be achieved. We assess pessimism/optimism (via the proxy of years-to-eradication), broken down by area of expertise, and adjust for selection bias. Our results show a disconnect between optimistic public institutional discourse and pessimistic private opinion, suggesting either (a) the necessity of a realistic accounting of eradication's low short-term likelihood in health planning and financing, or (b) the need for a more compelling case regarding eradication's feasibility to be made to the research community.
\end{changemargin}
\end{center}

@@

The WHO Global Malaria Programme (GMP) has acknowledged that it "needs to take an official position on how and under what timeline malaria eradication could be achieved" [@WHO2015]. Such a position could inform policy, and plays a crucial role in the economic analysis of the expected value of malaria control interventions.

@@

A great deal of previous research already covers the the cost per case prevented [@Sicuri2011] [@Silumbe2015] [@Btto2016 [@IlungaIlunga2014] [@Dalaba2015]. Likewise, literature exists which could serve as a model for quantifying the location-specific opportunity costs associated with funneling funds towards malaria eradication [@Stuckey2014] [@White2011] [@Korenromp2012]. The correct discount rate for estimating the value of future lives saved is more of a philosophical question than an economic one. This leaves only the probability and time-frame to eradication, questions which have been addressed anecdotally, but never answered quantitatively.

@@

The objective of this study is to help guide the optimal distribution of health resources by gauging (expert) opinion about, estimating the likelihood of, and quantifying the time frame to malaria eradication through a systematic survey of malaria research professionals from a wide array of academic disciplines.

@@

Our study population included all first authors (with available email addresses) returned in a PubMed search for the term "malaria" from January 1, 2010 through December 20, 2016. Personalized emails addressing the author by name and mentioning the relevant paper were sent to each of the `r nrow(email_df)` authors during the period from December 20, 2016 through January 2, 2017. Researchers were invited to participate by clicking a link to the survey form. Participants were informed directly on the survey form of the study's purpose, that responses were not guarnteed to be anonymous, and where to find further details on the study.

The survey was simple, consisting of only name, email, and four content-related fields along with a "general comments" section. The survey was administered and data were collected through Google Drive. The original survey is viewable at [https://goo.gl/forms/IroAEooDuJ6KM5Ho2](https://goo.gl/forms/IroAEooDuJ6KM5Ho2).

@@

Qualitative analysis of free text comments was carried out using thematic content analysis, following a grounded theory approach [@Markey_2014]. Participants were invited to provide "any general comments on the timeline and likelihood of global eradication".  Thematic content analysis [@Vaismoradi_2016] was employed to code responses following the 6-phase approach laid out by Braun and Clarke [@Braun2006]. The approach was inductive, used open-coding to identify latent themes, and underwent several iterations in which codes were modified, discarded and created. Using the RQDA software to assist in data management and theme coding [@Ronggui2017], four subject themes were identified. Comments were additionally coded as either descriptive (comments pertaining to the "problem" of malaria eradication) or presciptive (pertaining to potential "solutions" for eradicating malaria). Finally, free-text comments were scored for overall sentiment polarity [@sentimentr2017].

@@

@@

A total of `r length(which(df$responded))` researchers participated in the survey from the `r nrow(df)` invitations sent (participation rate of `r round(length(which(df$responded)) / nrow(df) * 100, 2)`%). Areas of expertise were non-exclusive and self-described, with participants having the option to choose from up to 3 of 10 checkboxes\footnote{Responses categories were Anthropology, Biochemistry, Bioinformatics, Biology, Chemistry, Clinical medicine, Drug discovery, Ecology, Economics, Entomology, Epidemiology, Geography, GIS, History, Immunology, Infectious disease, IT, Malaria, Medical entomology, Medicinal chemistry, Microbiology, Parasitology, Pharmacology, Pharmacy, Political science, Public health, Statistics, Vector biology, Vector control, and Virology.}, or to write in one or more "other" areas of expertise. `r sum(x$n[1:3])` (`r round(sum(x$p[1:3]), digits = 1)`%) participants declared three or fewer areas of expertise.

@@

Respondents were qualitatively different from non-respondents. Importantly, the average number of total author-specific citations was `r round(mean(df$n_citations_total[df$responded]), digits = 2)` among respondents, but `r round(mean(df$n_citations_total[!df$responded]), digits = 2)` among non-respondents. This suggests either a tendency for more senior or more cited researchers not to respond. When examining the number of average citations per article, the difference between respondents and non-respondents was slightly less: `r round(median(df$n_citations_total[df$responded]), digits = 2)` among respondents, and `r round(median(df$n_citations_total[!df$responded]), digits = 2)` among non-respondents. Males responded at a greater rate (`r round(mean(df$responded[df$sex == 'male'], na.rm = T) * 100, digits = 2)`) than females (`r round(mean(df$responded[df$sex == 'female'], na.rm = T) * 100, digits = 2)`) and those with Hispanic last names responded at a greater rate than those with last names of other ethnicities (see table).

@@

@@

# ggplot(data = results) + 
#   geom_violin(aes(x = bb,
#            y = ranking),
#            fill = 'orange',
#               alpha = 0.6) +
#   geom_jitter(aes(x = bb,
#            y = ranking),
#            color = 'black',
#            with = 0.4,
#            height = 0.4,
#               alpha = 0.3,
#               size = 1) +
#   theme_paper(base_size = 10) +
#   labs(x = 'WHO Building block',
#        y = 'Ranking',
#        title = 'Perceived areas needing attention',
#        subtitle = 'Small points: jittered individual responses; large triangles: averages') +
#   geom_point(data = results_agg,
#              aes(x = bb,
#                  y = avg),
#              size = 5,
#              pch = 17,
#              color = 'black',
#              alpha = 0.8) +
#   ylim(0, 6)
```

@@

1. *Innovation*: comments pertaining to new tools, drugs, methods, or strategies required to achieve eradication.
2. *Systematic challenges*: comments pertaining to political, social, environmental or logistical issues related to eradication.
3. *Poverty*: comments specifically related to the economic and socieconomic components of malaria eradication.
4. *Meta*: challenging the question of eradication itself, re-framing the goal, insisting on other priorities, etc.

@@

Many noted the need to "overcome the challenge of drug resistance". More than 10% of commenters noted the need for an effective vaccine. Genetic engineering was mentioned by several commenters as a promising means to achieve eradication quickly. Most comments coded as "innovatation" were presciptive in nature, often suggesting the nature of the needed innovation, with a heavy slant towards pharmaceutical options and vaccination.

**Systematic challenges** to malaria eradication were noted by the majority of commenters. In direct contrast to the previous comments, many emphasized that "we already have the tools to achieve  eradication" and that the only piece lacking was "robust health systems". Many commenters noted problems of coordination:

\begin{quote}
It will be very difficult to eradicate malaria... not because we don't have the technologies, which we already have... the problem is politics. Malaria doesn't stop in borders of a country and it would take a joint effort of a lot of political leaders to get programs in place to fight malaria. Unfortunately I don't see this happening anytime soon.
\end{quote}

Others echoed the sentiment, with many comments focusing on the need of strong surveillance and treatment delivery systems. Many commenters focused on the reasons for stagnating progress, such as "weak or failing health systems...due to political unwillingness or conflict". Others highlighted the inherent challenges in the epidemiology of malaria, such as the changing dynamics of malaria transmission, the resilience of the parasite and vector, climate change, and the inability to aim interventions accurately with an "ever-moving target". Comments coded as "systematic" tended to be descriptive and slightly more pessimistic than others.

**Poverty** was assigned its own category in the codification of free text responses - although it is best understood as a sub-category of "systemtatic challenges" - because of its ubiquity as a theme in comments. Many noted that malaria is a "disease of poverty", with "social injustice" as the root cause. Some made the sequential argument that "eradication of poverty" must preceed disease eradication. Along the same lines, one commenter wrote:

@@

Another noted that the survey "left off the list the most important factor - economic development". Many echoed the sentiment, stating that "with no economic development, you cannot have eradication" and that poverty is the "cause" of malaria. As with the "systematic challenges" of eradication, comments which were coded as poverty tended to be descriptive rather than prescriptive.

**Meta** arguments were rarer than the other categories, but cohesive and consistent. One commenter wrote that eradication is a "postwar" idea that developed from the "abandonment of a broad sociopolitical understanding of the causes of disease, and the emphasis on technological solutions." Many stated that global malaria eradication was simply not possible, and 2 argued that it may not be desirable or ethical. One stated that the concept was "absurd" and that "I'm not even sure why people talk about it". Another questioned the utility of discussing "eradiction" as a concept:

@@

Though results are not disclosed at the individual level, this survey was not entirely anonymous, suggesting that social desirability bias may have influenced responses, and that selection bias likely meant our pool of participants were substantially different from those who choose not to participate.

@@

This study included the corresponding authors of indexed journals. Though certainly a group with important knowledge related to malaria, this misses malaria control program employees, health agency workers, and other stakeholders. Their experiences and viewpoints may be different from those of academicians, and arguably more relevant.

@@

Whether the collective "wisdom" of researchers is worth adhering to or not is arguable. That said, to the extent that researchers' private opinions differ so greatly from those of global health leaders is itself a noteworthy result. If leaders' high levels of optimism are well-placed, this suggests that researchers are unduly pessimistic, and efforts should be made to convince them that eradication is more feasible than they realize. If, on the other hand, collective wisdom is a better forecasting agent than the proclamations and objectives of global health leaders, then it is important to recognize the sad truth that we are likely further from eradication than we would like to be. In this case, though the discourse on eradication can be motivating and inspiring, resources should be directed in a manner that reflects an intended outcome's likelihood of occurrence.

Our survey attempted to quantify the likelihood of and time frame to malaria eradication by gauging the collective opinion of malaria researchers. Our results suggest that eradication is unlikely in the short and medium terms, and that there exists a great deal of disagreement among researchers in regards to which areas require the most attention in order for eradication to occur. The implication of these results are three-fold: (1) that those working or investing in eradication-specific campaigns should factor in the low likelihood and long time frame when calculating those campaigns' expected value; (2) that champions of near-term eradication need to make a more compelling case to malaria researchers of eradication's feasibility, in order to better focus and inspire the latter; and (3) that discord among malaria researchers as to which areas need the most attention in order for eradication to occur, suggests a need for open discussion and concensus-making, so that resources and focus are directed appropriately.

>>>>

---
title: "Silent pessimists: researchers' perceptions of malaria eradication and implications for health spending"
bibliography: library.bib
output: 
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: preamble.sty
---

@@

Quantifying an event's probability and time frame is essential for calculating its expected value. In the case of global malaria eradication, uncertainty regarding feasibility makes it difficult for policymakers and public health practitioners to make fully informed decisions. The opportunity cost of investments in eradication-specific interventions can be high, particularly in contexts with other urgent health priorities. In a systemic survey of malaria researchers, we query perceptions regarding the likelihood and time frame of eradication, as well as the perceived chief areas needed for improvement in order for eradication to be achieved. We assess pessimism/optimism (via the proxy of years-to-eradication), broken down by area of expertise, and adjust for selection bias. Our results show a disconnect between optimistic public institutional discourse and pessimistic private opinion, suggesting either (a) the necessity of a realistic accounting of eradication's low short-term likelihood in health planning and financing, or (b) the need for a more compelling case regarding eradication's feasibility to be made to the research community.
\end{changemargin}
\end{center}

@@

International programes, such as the WHO Global Malaria Programme (GMP), have acknowledged the need "to take an official position on how and under what timeline malaria eradication could be achieved" [@WHO2015]. Such a position could inform policy, and plays a crucial role in the economic analysis of the expected value of malaria elimination initiatives.

@@

A great deal of previous research already covers the cost per case prevented [@Sicuri2011] [@Silumbe2015] [@Btto2016 [@IlungaIlunga2014] [@Dalaba2015]. Likewise, literature exists which could serve as a model for quantifying the location-specific opportunity costs associated with funneling funds towards malaria eradication [@Stuckey2014] [@White2011] [@Korenromp2012]. The correct discount rate for estimating the value of future lives saved is more of a philosophical question than an economic one. This leaves only the probability and time-frame to eradication, questions which have been addressed anecdotally, but never answered quantitatively.

@@

The objective of this study is to help guide the optimal distribution of health resources by gauging (expert) opinion about, estimating the likelihood of, and quantifying the time frame to malaria eradication through a systemic survey of malaria research professionals from a wide array of academic disciplines.

@@

Our study population included all first authors (with available email addresses) returned in a PubMed search for the term "malaria" from January 1, 2010 through December 20, 2016. Personalized emails addressing the author by name and mentioning the relevant paper were sent to each of the `r nrow(email_df)` authors during the period from December 20, 2016 through January 2, 2017. Researchers were invited to participate by clicking a link to the survey form. The survey was simple, consisting of only name, email, and four content-related fields along with a "general comments" section. The survey was administered and data were collected through Google Drive. The original survey is viewable at [https://goo.gl/forms/IroAEooDuJ6KM5Ho2](https://goo.gl/forms/IroAEooDuJ6KM5Ho2).

@@

@@

Qualitative analysis of free text comments was carried out using thematic content analysis, with inductive open coding [@Markey_2014]. Participants were invited to provide "any general comments on the timeline and likelihood of global eradication".  Thematic content analysis [@Vaismoradi_2016] was employed to code responses following the 6-phase approach laid out by Braun and Clarke [@Braun2006]. The approach was inductive, used open-coding to identify latent themes, and underwent several iterations in which codes were modified, discarded and created. Using the RQDA software to assist in data management and theme coding [@Ronggui2017], four subject themes were identified. Comments were additionally coded as either descriptive (comments pertaining to the "problem" of malaria eradication) or presciptive (pertaining to potential "solutions" for eradicating malaria). Finally, free-text comments were scored for overall sentiment polarity [@sentimentr2017].

@@

A total of `r length(which(df$responded))` researchers participated in the survey from the `r nrow(df)` invitations sent (participation rate of `r round(length(which(df$responded)) / nrow(df) * 100, 2)`%). Areas of expertise were non-exclusive and self-described, with participants having the option to choose from up to 3 of 10 checkboxes, or to write in one or more "other" areas of expertise.\footnote{Response categories were Anthropology, Biochemistry, Bioinformatics, Biology, Chemistry, Clinical medicine, Drug discovery, Ecology, Economics, Entomology, Epidemiology, Geography, GIS, History, Immunology, Infectious disease, IT, Malaria, Medical entomology, Medicinal chemistry, Microbiology, Parasitology, Pharmacology, Pharmacy, Political science, Public health, Statistics, Vector biology, Vector control, and Virology.} `r sum(x$n[1:3])` (`r round(sum(x$p[1:3]), digits = 1)`%) participants declared three or fewer areas of expertise.

@@

Respondents were qualitatively different from non-respondents. Importantly, the average number of total author-specific citations was `r round(mean(df$n_citations_total[df$responded]), digits = 2)` among respondents, but `r round(mean(df$n_citations_total[!df$responded]), digits = 2)` among non-respondents. This suggests either a tendency for more senior or impactful researchers not to respond. When examining the number of average citations per article, the difference between respondents remained: `r round(mean(df$n_citations_average[df$responded]), digits = 2)` among respondents, and `r round(mean(df$n_citations_average[!df$responded]), digits = 2)` among non-respondents, highlighting the greater impact of non-respondents relative to respondents. Males responded at a greater rate (`r round(mean(df$responded[df$sex == 'male'], na.rm = T) * 100, digits = 2)`) than females (`r round(mean(df$responded[df$sex == 'female'], na.rm = T) * 100, digits = 2)`) and those with Hispanic last names responded at a greater rate than those with last names of other ethnicities (see table).

@@

Of the below two sections only one will remain in the paper: either (i) de-biasing using weighted binomial regression or (ii) de-biasing using Heckman.

## De-biasing using weighted binomial regression

```{r}
# Estimate model for de-biasing sample selection without Heckman.
response_model <- glm(responded ~ bin_citations + sex, data = df, family = binomial('logit'))
# Use model to predict likelihood of response
df$likelihood_of_response <- predict(response_model, df, type = 'response')
# Use inverse of likelihood of response to assign weight for model
df$weight <- 1 - df$likelihood_of_response
```

## De-biasing using Heckman

@@

# ggplot(data = results) + 
#   geom_violin(aes(x = bb,
#            y = ranking),
#            fill = 'orange',
#               alpha = 0.6) +
#   geom_jitter(aes(x = bb,
#            y = ranking),
#            color = 'black',
#            with = 0.4,
#            height = 0.4,
#               alpha = 0.3,
#               size = 1) +
#   theme_paper(base_size = 10) +
#   labs(x = 'WHO Building block',
#        y = 'Ranking',
#        title = 'Perceived areas needing attention',
#        subtitle = 'Small points: jittered individual responses; large triangles: averages') +
#   geom_point(data = results_agg,
#              aes(x = bb,
#                  y = avg),
#              size = 5,
#              pch = 17,
#              color = 'black',
#              alpha = 0.8) +
#   ylim(0, 6)

# a <- data_frame(category = rep(c("Innovation",
#                                                    "Systemic Challenges",
#                                                    "Poverty",
#                                                    "Meta"),2),
#                 type = rep(c('Prescriptive', 'Descriptive'), each = 4),
#                 val = c(213,
#                         25,
#                         85,
#                         32,22,
#                         112,
#                         70,
#                         142))
# a$val <- ifelse(a$type == 'Prescriptive', a$val * -1, a$val)
# ggplot(data = a,
#        aes(x = category,
#            y = val)) +
#   geom_bar(alpha = 0.5,stat = 'identity', aes(fill = type)) +
#   coord_flip() +
#   geom_label(aes(label = abs(val))) +
#   theme(legend.position = 'bottom') +
#   scale_fill_manual(name = '',
#                     values = c('lightblue', 'darkorange')) +
#   theme_fivethirtyeight() +
#   ylim(-220, 150) +
#   labs(title = 'Classification of survey free responses',
#        subtitle = 'By thematic categorization and diagnostic')
```

@@

1. *Innovation*: comments pertaining to new tools, drugs, methods, or strategies required to achieve eradication.
2. *Systemic challenges*: comments pertaining to political, social, environmental or logistical issues related to eradication.
3. *Complexity*: comments which of focus on the multi-dimensional components of eradication.
4. *Tackling the fundamentals*: challenging the question of eradication itself, re-framing the goal, insisting on other priorities, etc.

@@

Many noted the need to "overcome the challenge of drug resistance". More than 10% of commenters noted the need for an effective vaccine. Genetic engineering was mentioned by several commenters as a promising means to achieve eradication quickly. Most comments coded as "innovation" were presciptive in nature, often suggesting the nature of the needed innovation, with a heavy slant towards pharmaceutical options and vaccination.

**Systemic challenges** to malaria eradication were noted by the majority of commenters. Comments in this category can be divided into roughly four sub-themes: (i) lack of coordination, (ii) lack of good surveillance and health services delivery, (iii) lack of political will and (iv) poverty. In direct contrast to the previous comments, many emphasized that "we already have the tools to achieve  eradication" and that the only piece lacking was "robust health systems". Many commenters noted problems of coordination, as illustrated in the below quote.

\begin{quote}
It will be very difficult to eradicate malaria... not because we don't have the technologies, which we already have... the problem is politics. Malaria doesn't stop in (sic) borders of a country and it would take a joint effort of a lot of political leaders to get programs in place to fight malaria. Unfortunately I don't see this happening anytime soon.
\end{quote}

Others echoed the sentiment, with many comments focusing on the need of strong surveillance and treatment delivery systems. Many commenters focused on other reasons for stagnating progress, such as "weak or failing health systems...due to political unwillingness or conflict".  Comments coded as "systemic" tended to be descriptive and slightly more pessimistic than others. Many noted that malaria is a "disease of poverty", with "social injustice" as the root cause. Some made the sequential argument that "eradication of poverty" must preceed disease eradication. Along the same lines, one commenter wrote:

@@

Another noted that the survey "left off the list the most important factor - economic development". Many echoed the sentiment, stating that "with no economic development, you cannot have eradication" and that poverty is the "cause" of malaria. As with the "systemic challenges" of eradication, comments which were coded as poverty tended to be descriptive rather than prescriptive.

**Complexity** was a relatively rare category (<20\% of all comments), those whose comments were coded as the "complexity" category were slightly more pessimistic than average in regards to the timeline and likelihood of eradication. Many commenters highlighted the inherent challenges in the epidemiology of malaria, such as the changing dynamics of malaria transmission, the resilience of the parasite and vector, climate change, and the inability to aim interventions accurately with an "ever-moving target". The potential for adaptation was highlighted in reference both to the mosquito as well as the parasite itself. Many comments addressed the fact that the conversation on eradication is largely taking place within the public health community, whereas the causes of malaria endemicity are largely orthogonal to public health interventions, "going beyond the health sector". Several commenters pointed out the multitude of prerequisite conditions for eradication to even be considered feasible:

\begin{quote}
To my mind, this question is highly dependent on background context, e.g. global political and economic dynamics, as well as international conflict. Complete global eradication is an extremely singular goal that requires a vast array of necessary conditions - if any of these fail, eradication will not be achieved.  
\end{quote}

**Tackling the fundamentals** was rarer than the other categories, but cohesive and consistent. One commenter wrote that eradication is a "postwar" idea that developed from the "abandonment of a broad sociopolitical understanding of the causes of disease, and the emphasis on technological solutions." Many stated that global malaria eradication was simply not possible, and two argued that it may not be desirable or ethical. One stated that the concept was "absurd" and that "I'm not even sure why people talk about it". Another questioned the utility of discussing "eradiction" as a concept:

@@

@@

This study included the corresponding authors of indexed journals. Though certainly a group with important knowledge related to malaria, this misses malaria control program employees, health agency workers, and other stakeholders. Their experiences and viewpoints may be different from those of academics, and arguably more relevant.

@@

Whether the collective "wisdom" of researchers is worth adhering to or not is arguable. That said, to the extent that researchers' private opinions differ so greatly from those of global health leaders is itself a noteworthy result. If leaders' high levels of optimism are well-placed, this suggests that researchers are unduly pessimistic, and efforts should be made to convince them that eradication is more feasible than they realize (so that efforts are directed accordingly). If, on the other hand, collective wisdom is a better forecasting agent than the states objectives of global health leaders, then it is important to recognize the sad truth that we are likely further from eradication than we would like to be. In this case, though the discourse on eradication can be motivating and inspiring, resources should be directed in a manner that reflects an intended outcome's likelihood of occurrence.

Our survey attempted to quantify the likelihood of and time frame to malaria eradication by gauging the collective opinion of malaria researchers. Our results suggest that eradication is unlikely in the short and medium terms, and that there exists a great deal of disagreement among researchers in regards to which areas require the most attention in order for eradication to occur. The causes for pessimism are diverse, but common themes were the need for innovation, systemic challenges, and the complexity of the disease and its transmission, along with a notable share of researchers who question the utility of the eradication concept itself. The implication of these results are three-fold: (1) that those working or investing in eradication-specific campaigns should factor in the low likelihood and long time frame when calculating those campaigns' expected value; (2) that champions of near-term eradication need to make a more compelling case to malaria researchers of eradication's feasibility, in order to better focus and inspire the latter; and (3) that discord among malaria researchers as to which areas need the most attention in order for eradication to occur, suggests a need for open discussion and concensus-making, so that resources and focus are directed appropriately.
